---
title: Some Base R Functions
date: "2024-11-26"
draft: true
category: "[code, r, base-r]"
engine: knitr
execute:
  echo: false
---

Let's go through a few `base R` functions that we may have underutilized.

::: callout-note
```{r}
base = installed.packages(.Library, priority = "base")[, "Package"]
high = installed.packages(.Library, priority = "high")[, "Package"]

base_pkgs = knitr::combine_words(base, before="`")
```

**Base R** refers to built-in packages that come with R[^almost]. Some of us don't probably use or even know exist some of these packages. According to the documentation (see [?installed.packages](https://search.r-project.org/R/refmans/utils/html/installed.packages.html)), Base R includes `{r} base_pkgs`.
:::

[^almost]: Packages that come with R includes more than base packages: `installed.packages(priority="high")` shows `base + recommended` packages.

### `base::file.path`

I see people use the [`{here}`](https://github.com/r-lib/here) package a lot, but it has not yet occurred to me that I needed it in my workflows when I already have `file.path`. While the goal of the package is to replace `setwd`, I often see people use it to construct file paths relative to the project directory like:

``` r
# Taken from the repo (see the readme)
path = here::here("inst", "demo-project", "data", "penguins.csv")
#> [1] "/home/kirill/git/R/here/inst/demo-project/data/penguins.csv"
write.csv(palmerpenguins::penguins, path)
```

If that is the sole use case, we can just use `file.path` provided we *setup our working directory properly*. Just replace `here::here` with `file.path` and we are good to go.

```{r}
#| echo: true
#| eval: false

path = file.path("inst", "demo-project", "data", "penguins.csv")
#> [1] "inst/demo-project/data/penguins.csv"
write.csv(palmerpenguins::penguins, path)
```
Notice that `file.path` provides the relative path, while `here::here` provides the (full) absolute path.

Of course, `{here}` is more powerful (has sophisticated project root resolving mechanisms, robust to in-session working dir changes (with setwd()), and [more](https://here.r-lib.org/articles/here.html)). But if you are using projects and work with (project) relative paths, `file.path` may do the job just fine. This feature of project-relative path construction of `{here}` (i.e., with `here::here`) is the comparable part to `file.path`.


### `file.stem` = `tools::file_path_sans_ext` + `tools::file_ext`

These functions allow for the manipulation of file paths, with `file_path_sans_ext` returning the file path without the extension and `file_ext` returning the file extension.

```{r}
#| echo: true
#| eval: false

file_path = "~/data/mtcars.csv"

# Extract the file path without the extension
tools::file_path_sans_ext(file_path)
#> [1] "~/data/mtcars"

# Extract the file extension
tools::file_ext(file_path)
#> [1] "csv"
```

At times, I need the stem of the file path to save a file with a different extension or to use the file path as an id in data frames when reading multiple files.

```{r}
#| eval: false

files = list.files("data", "*.csv", full.names = TRUE)
names(files) = tools::file_path_sans_ext(files) |> basename()

lapply(files, \(f) data.table::fread(f)) |>
  data.table::rbindlist()
```

I use the `tools::file_path_sans_ext` and `base::basename` combo a lot, so I put it into a function `file.stem` to use as a util in my projects.


### `base::sprintf`

This function serves as a workhorse for creating formatted strings. According to the [doc](https://search.r-project.org/R/refmans/base/html/sprintf.html), it is a wrapper for the C `sprintf` function, which suggests it is a common operation. It is very useful for creating consistent output in reports or visualizations.

You may have used `paste0` with some separators, and `round` to truncate numeric data, but `sprintf` is more powerful and flexible.

If you have done `paste0("a", "-", "b", "-", "c")`, the equivalent would be `sprintf('%s - %s - %s', "a", "b", "c")`.

The {glue} package is a popular and great alternative for formatting strings if you are not a dependency-averse person. FYI, glue has no dependencies.

```{r}
# create a string with placeholders for variables
my_string = "The value of x is %d and the value of y is %f"

# insert values into the placeholders
sprintf(my_string, 1, 2.5)
```


### `base::lengths`

This function returns the lengths of objects in a list, making it useful for quickly summarizing the structure of a data set.

```{r}
# create a list of vectors
my_list = list(c(1, 2, 3), c("a", "b", "c"), c(TRUE, TRUE, FALSE))

# calculate the lengths of each vector in the list
lengths(my_list)
```

### `rle`

This function computes the lengths and values of runs of equal values in a vector, making it useful for identifying patterns in data.

```{r}
# create a vector with repeated values
my_vector = c(1, 1, 1, 2, 2, 3, 3, 3, 3)

# calculate the lengths and values of runs of equal values in the vector
rle(my_vector)
```

### `ave`

This function calculates the mean of a vector, with the option to specify grouping variables for calculating means by group.

```{r}
# create a data frame with two variables
my_data = data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c("a", "b", "c", "d", "e")
)

# calculate the mean of variable "x"
ave(my_data$x)

# calculate the mean of variable "x" by the levels of variable "y"
ave(my_data$x, my_data$y)
```



### `sweep`

This function allows you to apply a function to a matrix or vector by a margin, making it useful for quickly performing calculations across rows or columns.

```{r}
# create a matrix
my_matrix = matrix(1:9, nrow = 3)

# apply the sum function across the rows of the matrix
sweep(my_matrix, 1, colSums(my_matrix), "+")
```

### `Map`

This function applies a function to each element of a list, returning a list of results.

```{r}
# create a list of vectors
my_list = list(r=rnorm(10), y=runif(10))

# apply the sum function to each vector in the list
Map(sum, my_list)
```

### `replace`

This function allows for the replacement of values in a vector or data frame, making it useful for data cleaning and manipulation.

```{r}
# create a vector with some missing values
my_vector = c(1, NA, 2, NA, 3)

# replace the missing values with 0
replace(my_vector, is.na(my_vector), 0)
```

### `within`

This function allows for the evaluation of an expression within a specified environment, making it useful for controlling variable scope and accessing objects in different environments.

```{r}
# create a data frame
my_data = data.frame(
  x = c(1, 2, 3, 4),
  y = c("a", "b", "c", "d")
)

within(my_data, mean(x))
```

### `strrep`

is used to repeat a string a specified number of times. Here is an example of how to use it:

```{r}
# Repeat the string "hello" 5 times
strrep("hello", 5)
```

This function can be useful for creating strings of repeated characters or words, or for padding strings with a certain character or word. It can also be useful for creating strings of a certain length by repeating a given string until it reaches the desired length.

```{r}
# Pad the string "hello" with underscores until it is 10 characters long
padded_string = paste("hello", strrep("_", 10 - nchar("hello")))
```
